---
title: "04. Generate peak-to-gene links"
output: html_notebook
---

# Identify putative cis-regulatory elements of the embryonic mouse cerebellum
Here, we used the ArchR addPeak2GeneLinks function to identify peak-to-gene links with the default settings except for changing the maximal distance between the peak and gene to 500 kb. We reasoned that some peak-to-gene links might be specific to each embryonic stage. Therefore, we subset the data according to stage, and iterated the analysis and identified additional peak-to-gene links that were specific to each stage.

```{r message=F}
library(ArchR);library(cowplot);library(Seurat);library(Signac)
set.seed(123)
addArchRThreads(threads = 20)
proj <- loadArchRProject(path = "E12_14")
```

# Split data according to stages
```{r}
# subseting E12.5 data
idxSample <- BiocGenerics::which(proj$Sample %in% "E12")
cellsSample <- proj$cellNames[idxSample]

proj.12 <- subsetArchRProject(
  ArchRProj = proj,
  cells = cellsSample,
  outputDirectory = "E12_subset",
  dropCells = TRUE,
  logFile = NULL,
  threads = getArchRThreads()
)
  
p12 <- plotEmbedding(ArchRProj = proj.12, colorBy = "cellColData", name = "cellType", embedding = "UMAP")
plotPDF(p12, name = "E12_subset.pdf", ArchRProj = proj, addDOC = FALSE, width = 5, height = 5)

# subseting E13.5 data
idxSample <- BiocGenerics::which(proj$Sample %in% "E13")
cellsSample <- proj$cellNames[idxSample]
proj.13 <- subsetArchRProject(
  ArchRProj = proj,
  cells = cellsSample,
  outputDirectory = "E13_subset",
  dropCells = TRUE,
  logFile = NULL,
  threads = getArchRThreads()
)

p13 <- plotEmbedding(ArchRProj = proj.13, colorBy = "cellColData", name = "cellType", embedding = "UMAP")
plotPDF(p13, name = "E13_subset.pdf", ArchRProj = proj, addDOC = FALSE, width = 5, height = 5)

# subseting E14.5 data
idxSample <- BiocGenerics::which(proj$Sample %in% "E14")
cellsSample <- proj$cellNames[idxSample]
proj.14 <- subsetArchRProject(
  ArchRProj = proj,
  cells = cellsSample,
  outputDirectory = "E14_subset",
  dropCells = TRUE,
  logFile = NULL,
  threads = getArchRThreads()
)

p14 <- plotEmbedding(ArchRProj = proj.14, colorBy = "cellColData", name = "cellType", embedding = "UMAP")
plotPDF(p14, name = "E14_subset.pdf", ArchRProj = proj, addDOC = FALSE, width = 5, height = 5)

pdf("E12_14/Plots/All_subsets1.pdf", h =5, w =15)
cowplot::plot_grid(p12,p13,p14,ncol = 3)
dev.off()
```

# Generate peak-to-gene links in each dataset
```{r}
list = c("E12_14","E12_subset","E13_subset","E14_subset")

geneSet <- readRDS("geneSet.rds")
for (ArchR_proj in list){
  proj <- loadArchRProject(path = ArchR_proj)
  
  # Identify peak-to-gene links 
  proj <- addPeak2GeneLinks(
    ArchRProj = proj,
    reducedDims = "IterativeLSI",
    useMatrix = "GeneIntegrationMatrix",
    dimsToUse = 1:30,
    scaleDims = NULL,
    corCutOff = 0.75,
    cellsToUse = NULL,
    k = 100,
    knnIteration = 500,
    overlapCutoff = 0.8,
    maxDist = 250000,
    scaleTo = 10^4,
    log2Norm = TRUE,
    predictionCutoff = 0.4,
    addEmpiricalPval = FALSE,
    seed = 1,
    threads = max(floor(getArchRThreads()/2), 1),
    verbose = TRUE,
    logFile = createLogFile("addPeak2GeneLinks")
  )
  
  p2g <- getPeak2GeneLinks(
    ArchRProj = proj,
    corCutOff = 0.45,
    resolution = 1,
    returnLoops = FALSE
  )
  
  p2g.peaks <- proj@peakSet[p2g$idxATAC,]
  p2g.peaks$peaks <- GRangesToString(grange = p2g.peaks, sep = c(":", "-"))
  p2g.peaks$linkedGene <- geneSet$name[p2g$idxRNA]
  p2g.peaks$Correlation <- p2g$Correlation
  p2g.peaks$FDR <- p2g$FDR
  p2g.peaks$VarQATAC <- p2g$VarQATAC
  p2g.peaks$VarQRNA <- p2g$VarQRNA
  p2g.peaks
  
  p2g.db <- data.frame(p2g.peaks)
  p2g.db$nearest_TSS <- geneSet$start[match(p2g.db$nearestGene, geneSet$name)]
  p2g.db$linked_TSS <- geneSet$start[match(p2g.db$linkedGene, geneSet$name)]
  p2g.db <- dplyr::mutate(p2g.db,distToTSS_near = abs(nearest_TSS-abs(start+end)/2),
                   distToTSS_linked = abs(linked_TSS-abs(start+end)/2))

  saveRDS(p2g.db, paste0(ArchR_proj,"/results/peak2gene.rds"))
}
```

